---
title: "Homework 6 - Taylor Kielczewski"
format: html
editor: visual
---

# **Complex Linear Models**

**github**

https://github.com/taylor-kielczewski/Biostatistics

## **1. Complex Linear Models and their Design Matrices**

```{r}
# loading libraries & data

library(dplyr)
library(palmerpenguins)
data(penguins)
library(performance)
library(knitr)
library(ggplot2)
library(readr)
library(tidyverse)
library(broom)
library(visreg)
library(modelr)

penguins <- penguins |> filter(!is.na(sex))

```

## **1a.**

```{r}
pen_mod <- lm(body_mass_g ~ flipper_length_mm + sex, data = penguins)

pen_mod

```

## **1b**

| intercept | flipper_length_mm | sexmale |
|-----------|-------------------|---------|
| 1         | 181               | 1       |
| 1         | 186               | 0       |
| 1         | 195               | 0       |

```{r}
  
model.matrix(body_mass_g ~ flipper_length_mm + sex, data = penguins)[1:3, ]


```

## **1c**

```{r}

contrasts(penguins$sex)
contrasts(penguins$species) <- contr.sum(3)

pen_mod2 <- lm(body_mass_g ~ flipper_length_mm + sex + species, data = penguins)


pen_mod2
```

```{r}
X <- model.matrix(body_mass_g ~ flipper_length_mm + sex + species, data = penguins)

rows <- c(1, 2, 200, 201, 300, 301)

X[rows,]
```

| intercept | flipper_length_mm | sexmale | species1 | species2 |
|-----------|-------------------|---------|----------|----------|
| 1         | 181               | 1       | 1        | 0        |
| 1         | 186               | 0       | 1        | 0        |
| 1         | 217               | 0       | -1       | -1       |
| 1         | 220               | 1       | -1       | -1       |
| 1         | 195               | 1       | 0        | 1        |
| 1         | 199               | 0       | 0        | 1        |

## **1d**

```{r}

pen_mod3 <- lm(body_mass_g ~ flipper_length_mm*sex + species, data = penguins)

pen_mod3

```

```{r}
Y <- model.matrix(body_mass_g ~ flipper_length_mm*sex + species, data = penguins)



Y[rows,]
```

| intercept | flipper_length_mm | sexmale | species1 | species2 | flipper_length_mm:sexmale |
|------------|------------|------------|------------|------------|------------|
| 1 | 181 | 1 | 1 | 0 | 181 |
| 1 | 186 | 0 | 1 | 0 | 0 |
| 1 | 217 | 0 | -1 | -1 | 0 |
| 1 | 220 | 1 | -1 | -1 | 220 |
| 1 | 195 | 1 | 0 | 1 | 195 |
| 1 | 199 | 0 | 0 | 1 | 0 |

## ** 2. Three variations on a theme **

```{r}
# loading libraries and data

pacman::p_load(DAAG)
data(leaftemp)


```

```{r}
lm1 <- lm(tempDiff ~ vapPress, data = leaftemp) 
  plot(lm1)

lm2 <- lm(tempDiff ~ vapPress + CO2level, data= leaftemp)
  plot(lm2)

lm3 <- lm(tempDiff ~ vapPress*CO2level, data= leaftemp) 
  plot(lm3)
```

### **2b.**

```{r}
check_model(lm1)
```

```{r}
check_model(lm2)
```

```{r}
check_model(lm3)
```

lm1 and lm2 pass checks of the assumptions. lm3 passes all but the collinearity check.

## **2c**

$$ tempdiff_i = \beta_0 + (\beta_1 \times vapPress_i) + (\beta_2 \times Elevated_i) + (\beta_3 \times High_i) + \epsilon_i $$

## **2d**

```{r}
leaftemp$pred_lm1 <- predict(lm1)
leaftemp$pred_lm2 <- predict(lm2)
leaftemp$pred_lm3 <- predict(lm3)
```

```{r}
lm1_plot <- ggplot(data = leaftemp, aes(y = tempDiff, x = vapPress)) + 
  geom_jitter(alpha = .5) +
  geom_line(aes(y = pred_lm1), color = "blueviolet")
 
lm1_plot + labs( title = "Simple Linear Regression")
```

```{r}
lm2_plot <- ggplot(leaftemp, aes(x = vapPress, y = tempDiff, color = CO2level)) +
  geom_point(alpha = .8) +
  geom_line(aes(y = pred_lm2, group = CO2level)) +
  scale_color_viridis_d()
 
lm2_plot + labs( title = "ANCOVA")
```

```{r}
lm3_plot <- ggplot(data = leaftemp, aes(x=vapPress, y = tempDiff, color = CO2level)) +
  geom_point(alpha = .7) +
  geom_line(aes(y=pred_lm3)) +
  scale_color_brewer(palette = "Dark2")

lm3_plot + labs(title = "Interaction")
```

# **Part 3: Interactions with Continuous Variables**

```{r}
# Loading Data
detritivore <- read_csv("data/c_burial_sims.csv")

tibble(detritivore)
```

## **3a**

```{r}
det_lm <- lm(log(net_carbon_sequestration) ~ log(microbial_abundance * detritivore_biomass * sink_rate), data = detritivore)

det_lm |> plot()
```

```{r}
check_model(det_lm)
```

Homogeneity of variance is still bad - we have to center the model.

```{r}
data <- detritivore %>%
  mutate(
    sink_c = sink_rate - mean(sink_rate, na.rm = TRUE),
    microbe_c = microbial_abundance - mean(microbial_abundance, na.rm = TRUE),
    detriti_c = detritivore_biomass - mean(detritivore_biomass, na.rm = TRUE)
  )
```

```{r}
centered_model <- lm(net_carbon_sequestration ~ sink_c * microbe_c * detriti_c, data = data)

centered_model |> plot()
```

```{r}
check_model(centered_model)
```

## **3b**

```{r}
summary(centered_model)
```

$\beta_0$ = intercept, the expected sequestration when all predictors are at their means

$\beta_1$ = effects of sinking rate when other predictors are at their mean 
$\beta_2$ = effects of microbial abundance when other predictors are at their mean 
$\beta_3$ = effects of detritivore biomass when other predictors are at their mean

$\beta_4$ = combined effects of sinking rate and microbial abundance $\beta_5$ = combined effects of sinking rate and detritivore biomass $\beta_6$ = combined effects of detritivore biomass and microbial abundance

$\beta_7$ = combined effects of all three predictors

## **3c**

```{r}
det_predict <- augment(centered_model, interval="confidence")
det_predict
```
```{r}
det_predict <- det_predict |> 
  mutate(detri_facet = cut(detriti_c,
  breaks = quantile(detriti_c, probs = c(0, 1/3, 2/3, 1)),
                    labels = c("Low", "Medium", "High"),
                    include.lowest = TRUE))
```

```{r}
ggplot(det_predict, aes(x = sink_c, y = .fitted, color = microbe_c)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~detri_facet) +
  scale_color_viridis_c() +
  labs(
    title = "Net Carbon Sequestration",
    x = "Sink Rate",
    y = "Net Carbon Sequestration",
    color = "Microbial Abundance"
  ) +
  theme_minimal()
  
```
# **Meta Questions**

## Question 1. 

Maybe interactions with discrete variables? 

## Question 2. 

I think the transformation and visualization is the most tricky for me. I am having trouble figuring out how to handle the data and regression, and how to plot it so all of the information is available. 

## Question 3.

I think I will use LM with multiple predictors in response to my kelp mapping project. 

## Question 4.

I am a bit lost still -- I don't know when/what transformation of the data will allow it to keep it's integrity, and what would ruin the model.

## Question 5.

4-5 hours

## Question 6.

Sufficient. I feel okay on parts 1 & 2, but the viz on part 3 really challenged me. 

