---
title: "t_kielczewski_hw1"
format: html
editor: visual
---

# Homework 1

### by Taylor Kielczewski

### BIOL607

```{r}
data(quakes)

str(quakes)

```

```{r}
summary(quakes)
```

```{r}
quakes$long
```

```{r}
unique(quakes$stations)
```

```{r}
length(unique(quakes$stations))
```

There are 102 unique stations.

```{r}
range(quakes$depth)
```

The range of earthquake depths are between 40-680

```{r}
max(quakes$mag)
quakes[quakes$mag == 6.4, ]
```

The largest magnitude earthquake was measured at a 6.4. It was measured at station #122, which is at (-15.56, 167.62).

## A Little LaTeX

$$
x = \mu 
$$

One dollar sign creates an inline equation, like $Y_i = 3X_i$

Two dollar signs creates a display equation, like this:

$$
Y_i = 3X_i 
$$

### Superscripts

$$
a^2 + b^2 = c^2 
$$

### Normal Distribution

$$ 
y_i \sim N(\hat{y_i},\sigma^2)
$$

### Meta Questions

1.  It was less difficult than I expected. I think that by familiarizing myself with a little bit of code during January break helped whittle away some of the anxiety regarding R. The LaTeX, was the most difficult/tedious part for me.

2.  It was tricky, but the links that were provided were very helpful. I think that it will be useful outside of class for writing about statistical analysis.

3.  I think it took me about an hour (but I broke it up into several sessions)

4.  Sufficient. I felt confident with all of the questions, but had a little trouble with finding the location of the max magnitude earthquake. That problem required a little extra research and troubleshooting. 
