---
title: "02_Kielczewski_Taylor"
format: html
editor: visual
---

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(forcats)
library(paletteer)
library(gganimate)
library(ggridges)
library(ggthemes)
library(colorfindr)

theme_set(theme_bw(base_size=12))

ice <- read_csv("http://biol607.github.io/homework/data/NH_seaice_extent_monthly_1978_2016.csv") %>%
  mutate(Month_Name = factor(Month_Name),
         Month_Name = fct_reorder(Month_Name, Month))
```

### Box Plot

```{r}
library(ggplot2)
library(RColorBrewer)
library(tidyverse)

ggplot(data = ice, aes(x= Month_Name, y = Extent, fill = Month_Name,
                      )) +
  geom_boxplot( alpha = .75, show.legend = FALSE) + 
  labs( 
    title = "Variability in Sea Ice",
    x = "Month",
    y = "Extent of Ice",
    caption = "Data from Dave Curran",
    subtitle = "From 1978 - 2016 ") + 
 scale_fill_brewer(palette = "Paired") + 
  theme(
    plot.title = element_text(face = "bold"))
```

### Density Plot

```{r}
ggplot(data = ice, aes(x = Extent, y = Month_Name, fill = Month_Name)) +
  geom_density_ridges2(alpha = .85, show.legend = FALSE) + 
  labs( 
    title = "Variability in Sea Ice",
    x = "Extent of Ice",
    y = "Month",
    caption = "Data from Dave Curran",
    subtitle = "From 1978 - 2016 ") +
  scale_fill_brewer(palette = "Paired") + 
  theme(
    plot.title = element_text(face = "bold"))
```

### Faceted Line Plot

```{r}
ggplot(data = ice, aes(x = Year, y = Extent)) + 
  geom_line(linewidth = .5) +
  geom_smooth(method = "glm")+
  facet_wrap(vars(cut_interval(Month, n = 4)),
             labeller = as_labeller(c(
               "[1,3.75]" = "Winter", 
               "(3.75,6.5]" = "Spring", 
               "(6.5,9.25]" = "Summer", 
               "(9.25,12]" = "Fall"
             ))) +
  theme(strip.text = element_text(size = 16)) +
  labs(y = "Extent of Ice") 
```

### Making A Color Palette

#### I spent way too much time on this

```{r}
get_colors("https://encrypted-tbn0.gstatic.com/images?q=tbn:ANd9GcR0HAsDmyzatPtFBdywcOnS11wtiOIufZk-gA&s.jpg")  %>% 
  make_palette(n = 39)

my_palette5 <- c("#031927", "#0B2E41", "#0C4B74", "#033035", "#163A50", "#004157", "#04566C", "#1879A3", "#094871", "#3E7687", "#446E7C", 
                "#285E8A", "#2A77AB", "#1F537B", "#31759C", "#3F879F", "#53838D", "#61A483", "#5098A7", "#61A483", "#2C7353", "#298E64", 
                "#6CC686", "#28A5B7", "#82CD78", "#68D9A5", "#76DBA1", "#9EEDAB", "#99E9AE", "#86B5CF", "#7F8F9E", "#2DBFAA", "#4DBAC1", 
                "#33B585", "#A5F0AF", "#82CD78", "#9EEDAB", "#76DBA1", "#86B5CF")

```

### Line Plot Pre-Animation

```{r}

ggplot(data = ice, aes(x = Month_Name, y = Extent, color=as.factor(Year), group = Year)) +
  geom_line(show.legend = FALSE) +
  scale_color_manual(values = my_palette5) +
  labs( 
    title = "Variability in Sea Ice",
    x = "Extent of Ice",
    y = "Month",
    caption = "Data from Dave Curran",
    subtitle = "From 1978 - 2016 ") + 
  theme(
    plot.title = element_text(face = "bold")) 
```

### Animated!

```{r}


a <- ggplot(data = ice, aes(x = Month_Name, y = Extent, color = as.factor(Year), group = Year)) +
  geom_line(show.legend = FALSE) +
  scale_color_manual(values = my_palette5) +
  labs(title = "Variability in Sea Ice", subtitle = "Year: {round(frame_along)}", x = 'Month', y = 'Extent of Ice', color = 'Year', caption = "Data from Dave Curran") +
  theme_stata(base_family = "Times", base_size = 16) +
  transition_states(Year, transition_length = 2, state_length = 1) +
  enter_appear()+
  transition_reveal(along = Year)+ 
  ease_aes('linear') 

a
```

### Being Impressive

```{r}

ggplot(data = ice, aes(x = Year, y = Extent, color = Month_Name, fill = Month_Name))+
  geom_point(shape = 21) +
  geom_line(linewidth = 1.25)+
  theme_bw() +
  theme(
    text = element_text(family = "Optima", size = 16),
    plot.title = element_text(face = "bold", size = 28), panel.background = element_rect(fill = "grey99"), 
    plot.background = element_rect(fill = "honeydew2"), axis.title.x = element_text(size = 20), axis.title.y =element_text(size = 20)) +
  scale_colour_paletteer_d("ggsci::flattastic_flatui") +
  labs(title = "Sea Ice Variability", subtitle = "from 1976 - 2016", x = "Year", y = "Extent of Sea Ice", color = "Month", fill = "Month") +
  transition_states(Year, transition_length = 2, state_length = 1) +
  enter_appear()+
  transition_reveal(along = Year)+ 
  ease_aes('linear') +
  geom_segment(
    aes(x = Year, xend = Year, y = max(Extent) * 1.05, yend = Extent),  
    color = "grey25", linewidth = .7 , linetype = "solid", alpha = .3) +
  geom_text(
    aes(label = as.character(Year), x = Year, y = max(Extent)), 
    color = "grey25",alpha = .5, size = 5, fontface = "bold", vjust = -.5 , hjust = -.5)

```

## Meta 1

I am feeling a lot more confident about visualizing figures. ggplot makes it incredibly easy to compare different variables against each other quickly, and see trends/pattnerns. I am still grappling with the grammar of the language, but it's getting a bit more intuitive!

## Meta 2

My favorite thing is playing with all of the different aesthetic features to personalize each plot.

## Meta 3

I struggled with this assignment, but I feel like I learned a lot and am feeling more confident about plot-making in the future.

## Meta 4

Although I struggled with this assignment, I put in a lot of time and effort, and learned a lot about coding / troubleshooting. I would give myself a 'strong' rating for this assignment.
